---
title: "Market Analysis of Bird-Based Tourism Using ebird Database"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = F)
knitr::opts_knit$set(root.dir = "D:/ebird/data/birding_market/")
```
####eBird is an online database of bird observation trips (checklists) providing scientists and naturalists with information about bird distribution and abundance.

###Questions:
###1. Which metric can we use to measure the popularity of bird watching in different counties?
###2. How can we design successful birding tours?
###3. What demographic/ economic factors contribute to the popularity of bird watching?

```{r lib, message=F}
# Import libraries. Read files.
library(auk)
library(tidycensus)
library(tidyverse)
library(readxl)
library(urbnmapr)
library(gridExtra)
library(ggplot2)
library(corrplot)
library(modelr)
library(ranger)
library(gbm)
library(caret)
library(pdp)
library(lubridate)
library(ggmap)
library(gridExtra)
```

###1. Which metric can we use to measure the popularity of bird watching in different counties?
####1.1 Get the dataset
```{r,eval = FALSE}
#read and filter checklists for 2018-2020 https://ebird.org/science/download-ebird-data-products
checklists_filters <- auk_sampling('F:/ebd_sampling_relMar-2020/ebd_sampling_relMar-2020.txt', sep = "\t") %>% 
                      auk_date(date = c("2018-01-01", "2019-12-31")) %>% auk_country("United States") %>% auk_filter(file = 'filtered_checklist.txt')
checklists_raw <- read_sampling(checklists_filters,unique = TRUE)

checklists_raw <- checklists_raw %>% 
  select(checklist_id,state,county,county_code,observation_date,time_observations_started)
 
#get county level population, gender, age, race, citizen, income from US Census Bureau  https://www.census.gov/data/developers/data-sets/ACS-supplemental-data.html
census_raw <- get_acs(geography = "county", 
              variables = c(population = "B01003_001", male='B01001_002',female='B01001_026',median_age='B01002_001',
                            white='B02001_002',black='B02001_003',native='B02001_004',asian='B02001_005',pacific='B02001_006'                                                           ,nonecitizen='B05001_006',incomepercapita='B19301_001'), 
                            year = 2018, output = "wide") %>% select(-ends_with("M"))

```

```{r, include=FALSE}
checklists_raw <- readRDS("D:/ebird/data/birding_market/import_rawdata/checklists_raw.rds")
census_raw <- readRDS("D:/ebird/data/birding_market/import_rawdata/census_raw.rds")
```


####1.2 Data cleaning, preprocessing
```{r}
checklist_county_num <- checklists_raw %>% 
  group_by(state,county,county_code) %>% count(name='checklist_n') %>% ungroup() %>%  drop_na()

#Feature transformation
census_raw <- census_raw %>% 
  mutate(male=maleE/populationE, female=femaleE/populationE, white=whiteE/populationE,     black=blackE/populationE,native=nativeE/populationE,asian=asianE/populationE, pacific=pacificE/populationE,otherrace=1-white-black-native-asian-pacific,citizen=1-nonecitizenE/populationE) %>% 
  select(-(maleE:femaleE),-(whiteE:nonecitizenE))

#Employment statistics from Department of Labor, Education level from USDA
unemployment_raw <- read_xlsx('D:/ebird/data/birding_market/bls_laucnty18_labor.xlsx') %>% 
  unite("FIPS", c('State FIPS','county FIPS'),sep = '') %>%
  mutate(unemployment_p=`Unemployment rate(%)`) %>% 
  select(-'Unemployment rate(%)')

#https://www.ers.usda.gov/data-products/county-level-data-sets/
education_raw <- read_xls('D:/ebird/data/birding_market/Education_USDA.xls') %>% 
  mutate(FIPS=`FIPS Code`) %>% select(-`FIPS Code`)

census_raw <- census_raw %>% 
  inner_join(unemployment_raw,by = c("GEOID" = "FIPS")) %>% 
  inner_join(education_raw,by = c("GEOID" = "FIPS")) %>%
  mutate(county_fips=substr(GEOID, start = 3, stop = 5)) %>%
  separate(NAME,c(NA,'state'),sep=', ') %>%
            select(-c(County,'Area name'))

#https://github.com/birderboone/birderboone.github.io/blob/master/ebirdSpeciesModeled.csv
species_n <- read.csv('D:/ebird/data/birding_market/ebirdSpecies.csv')

checklist_county_num <- checklist_county_num %>% separate(county_code,c(NA,NA,'county_fips'),sep='-')

census_ebird_raw <- left_join(checklist_county_num,census_raw, by=c("state",'county_fips')) %>%
  mutate(county=tolower(county)) %>%
  left_join(species_n, by=c('State'='state','county'))
```


```{r 10, warning=F, message=F, echo=FALSE,cache=TRUE}
# Separating polyname into region and subregion
counties_fips <- maps::county.fips %>%
  separate(polyname, c("region","subregion"),',', FALSE) %>%
  separate(subregion, c('subregion','subsubregion'),':') %>%
  select(-subsubregion) %>%
  (function(df){df[!duplicated(df$fips), ]}) %>%
  mutate(fips = as.character(fips))

#draw the map shows the checklist number for each county
counties_sf <- get_urbn_map(map = "counties", sf = TRUE) %>% 
  separate('county_fips',c(NA,'county_fips'),sep=-3) %>%
  left_join(checklist_county_num, by = c("state_name"='state','county_fips')) %>%
  left_join(census_ebird_raw %>% 
              select(state,county_fips,populationE), by = c("state_name"='state','county_fips'))

counties_sf %>%
  ggplot() +
  geom_sf(mapping = aes(fill = log(checklist_n)),
          color = NA, size = 0.05) +
  labs(fill = "Log(Checklist number)")

```
```{r, message=F}
ggplot(census_ebird_raw %>% drop_na(), aes(x=log(populationE), y=log(checklist_n))) + 
  geom_point(shape=18, color="brown1")+
  geom_smooth(method=lm)+
  labs(title="Relationship between population 
       and checklist numbers in each county",
       x="Log(population)", y = "Log(checklist numbers)")+
  theme(plot.title = element_text(hjust = 0.5),text = element_text(size=15))
```

```{r, message=F}
ggplot(census_ebird_raw, aes(x=populationE)) + 
  geom_histogram()+
  xlim(0, 10000)+
  ylab("County count")+
  xlab("Population")

census_ebird_filtered <- census_ebird_raw %>% 
  filter(populationE>=2000)
```


####1.3 Use checklists per person as a measurement for the popularity of bird watching in each county
```{r}
census_ebird_filtered <- census_ebird_filtered %>% 
  mutate(checklist_pps=checklist_n/populationE) 

counties_sf %>% left_join(census_ebird_filtered %>% 
                            select(state,county_fips,checklist_pps), by = c("state_name"='state','county_fips')) %>%
  ggplot() +
  geom_sf(mapping = aes(fill = log(checklist_pps)),
          color = NA, size = 0.05) +
  labs(fill = "log(Checklist number per person)")

```


###2. Where should we start the business and how to design a successful product?
####2.1 Select the candidate county
```{r}
census_ebird_filtered_des <- readRDS("D:/ebird/data/birding_market/import_rawdata/census_ebird_filtered.rds")
census_ebird_filtered_des <- census_ebird_filtered_des %>% filter(populationE>=10000) %>%   #Population more than 10000
  select(state,county,checklist_n,populationE,totalSpSeen,checklist_pps) %>%
  arrange(desc(checklist_pps))
census_ebird_filtered_des
```

####2.2 Read the checklist information from Arizona Santa Cruz county
```{r,eval = FALSE}
checklists_sc <- checklists_raw %>% 
  filter(state=='Arizona',county=='Santa Cruz') %>% 
  select("checklist_id","observation_date","time_observations_started","duration_minutes", "effort_distance_km",'latitude','longitude') %>% 
  distinct() 
```

```{r, include=FALSE}
checklists_sc <- readRDS("D:/ebird/data/birding_market/import_rawdata/checklists_sc.rds")
```

```{r}
#Convert hh:mm:ss to decimal time
time_to_decimal <- function(x) {
  x <- hms(x, quiet = TRUE)
  hour(x) + minute(x) / 60 + second(x) / 3600
}

checklists_sc <- checklists_sc %>% 
  mutate(time_observations_started = time_to_decimal(time_observations_started),
  day_of_year = yday(observation_date), month_of_year=month(observation_date), weekday=wday(observation_date)) %>% 
  drop_na()
```

####2.3 People's favorite months and days for bird watching
```{r}
# summarize data by hourly bins
month_n <- checklists_sc %>% 
  group_by(month_of_year) %>% 
  summarise(n_checklists = n())

month_fig <- ggplot(month_n) +
  aes(x = month_of_year, y = n_checklists) +
  geom_line() +
  geom_point() +
  labs(x = "Month",
       y = "Checklists number (n)",
       title = "Checklists number for each month") +
  scale_x_continuous(breaks = 1:12,labels = c('January', 'February', 'March', 'April', 'May', 'June', 'July', 'August', 'September', 'October', 'November', 'December'))+
  theme(axis.text.x = element_text(angle = 30))
  
weekday_n <- checklists_sc %>% 
  group_by(weekday) %>% 
  summarise(n_checklists = n()) %>%
  ungroup() %>%
  mutate(weekday=weekday-1) %>%
  mutate(weekday=ifelse(weekday==0, 7,weekday))

weekday_fig <- ggplot(weekday_n) +
  aes(x = weekday, y = n_checklists) +
  geom_line() +
  geom_point() +
  labs(y = "Checklists number (n)",
       title = "Checklists number for each day") +
  scale_x_continuous(breaks = 1:7,labels=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")) +
    theme(axis.text.x = element_text(angle = 30))

# combine
grid.arrange(month_fig, weekday_fig, nrow=1)
```
```{r, message=F}
#Latitude and longitude of sf found through https://www.latlong.net/
sc_map <- get_stamenmap(bbox = c(left = -111.363808, top=31.727364, bottom=31.331352, right=-110.448482), zoom = 10)

checklist_spring <- checklists_sc %>% filter(month_of_year%in%c(3,4,5))

map_spring <- ggmap(sc_map) +
  #contour
  geom_density2d(data = checklist_spring, aes(x = longitude, y = latitude), size = 0.3)+ 
  #fill in the color according to density
  stat_density2d(data = checklist_spring, aes(x = longitude, y = latitude, fill = ..level.., alpha = ..level..), geom = "polygon") + 
  scale_fill_gradient(low = "green", high = "red") +
  scale_alpha(range = c(0.1, 1), guide = FALSE)+
  theme(plot.title = element_text(size = 14, face = "bold"))+
  labs(title = "Density of checklists in spring")

checklist_fall <- checklists_sc %>% filter(month_of_year%in%c(7,8,9))

map_fall <- ggmap(sc_map) +
  #contour
  geom_density2d(data = checklist_fall, aes(x = longitude, y = latitude), size = 0.3)+ 
  #fill in the color according to density
  stat_density2d(data = checklist_fall, aes(x = longitude, y = latitude, fill = ..level.., alpha = ..level..), geom = "polygon") + 
  scale_fill_gradient(low = "green", high = "red") +
  scale_alpha(range = c(0.1, 1), guide = FALSE) +
  theme(plot.title = element_text(size = 14, face = "bold"))+
  labs(title = "Density of checklists in fall" )

grid.arrange(map_spring, map_fall, nrow=1)
```

####2.4 People's preference for start time and trip duration
```{r}
breaks <- 0:24
labels <- breaks[-length(breaks)] + diff(breaks) / 2
starttime_n <- checklists_sc %>% 
  mutate(startt = cut(time_observations_started, 
                         breaks = breaks, 
                        labels = labels,
                         include.lowest = TRUE),
         startt = as.numeric(as.character(startt))) %>% 
  group_by(startt) %>% 
  drop_na() %>% 
  summarise(n_checklists = n())

starttime_fig <- ggplot(starttime_n) +
  aes(x = startt, y = n_checklists) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 0:24, limits = c(0, 24)) +
  labs(x = "Hours since midnight",
       y = "Checklists number (n)",
       title = "Bird watching start times")+
  theme(axis.text.x = element_text(angle = 45))


breaks <- 0:24
labels <- 1:24
duration_n <- checklists_sc %>%
  mutate(duration_h=duration_minutes/36) %>%
  filter(duration_h<=24) %>%
  mutate(duration = cut(duration_h, 
                        breaks = breaks, 
                        label = labels,
                        include.lowest = TRUE),
         duration = as.numeric(as.character(duration))) %>% 
  drop_na() %>% 
  group_by(duration) %>% 
  summarise(n_checklists = n())

# histogram
duration_fig <- ggplot(duration_n) +
  aes(x = duration, y = n_checklists) +
  geom_col(width = mean(diff(breaks)), color = "grey30", fill = "grey50") +
  scale_x_continuous(breaks = 1:24) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Duration (h)",
       y = "Checklists number (n)",
       title = "Bird watching durations distribution")+
  theme(axis.text.x = element_text(angle = 45))

# combine
grid.arrange(starttime_fig, duration_fig, nrow=1)
```

###3. Which demographic/ economic factors contribute to the popularity of bird watching? The demographic features for our potential customers (The demographic features of ebird users are not provided by the database).
####3.1 The Pearson correlation between checklist number per person and other features
```{r}
census_ebird_filtered <- census_ebird_filtered %>% 
  select(-c('county','State',"state","county_fips","checklist_n","GEOID","populationE","male","otherrace","Lesshighschool_p","highschool_p",  "college_p","checklists","modeledTotal","speciesDifference" )) %>% 
  filter(checklist_pps<=1) %>% 
  drop_na()

feature_cor=cor(census_ebird_filtered, method = "pearson")
corrplot(feature_cor, type="upper")
```


####3.2 Use linear regression to predict the popularity of bird watching in each county.
```{r}
normalize <- function(x) { return ((x - min(x)) / (max(x) - min(x)))}
census_ebird_nor <- census_ebird_filtered %>% mutate_at(vars(median_ageE:totalSpSeen), normalize)

#Use 80% samples to fit the model, use another 20% to test the model
set.seed(1) 
in.trn <- sample.int(nrow(census_ebird_nor), 0.8*nrow(census_ebird_nor)) 
census_ebird_nor_trn <- census_ebird_nor[ in.trn,]
census_ebird_nor_tst <- census_ebird_nor[-in.trn,]

linearMod <- lm(checklist_pps ~ ., data=census_ebird_nor_trn)  

linear_coeff <- as.data.frame(sort(linearMod$coefficients[-1]))
colnames(linear_coeff) <- 'coeff'
linear_coeff$feature <- rownames(linear_coeff)

ggplot(data=linear_coeff, aes(x=reorder(row.names(linear_coeff),coeff), y=coeff))+
  geom_bar(stat="identity")+
  theme(axis.text.x = element_text(angle = 30))+
  ylab("Linear coefficient")+
  xlab("Features")

error=matrix(0,nrow=2,ncol=2, dimnames = list(c('linear','GBM'), c('training error','test error')))
error[1,1]=sqrt(sum((linearMod$residuals)^2)/length(linearMod$residuals))

```

```{r}
tstPred <- predict(linearMod, census_ebird_nor_tst)

linearMod_result <- as.data.frame(cbind(actuals=census_ebird_nor_tst$checklist_pps, predicteds=tstPred))

rmse_linear <- sqrt(mean((linearMod_result$actuals - linearMod_result$predicteds)^2))
error[1,2]=rmse_linear
sprintf('The RMSE for linear model on test data is %f',rmse_linear)
```


####3.3 Use Gradient Boosting Machine to predict the popularity of bird watching in each county. Find the optimized parameters
```{r}
set.seed(66)
# train GBM model, find the range of iteration
gbm.fit <- gbm(
  formula = checklist_pps ~ .,
  distribution = "gaussian",
  data = census_ebird_nor_trn,
  n.trees = 10000,  #total number of train
  interaction.depth = 3,  #tree height, determine the number of splits
  shrinkage = 0.01,   #Learning rate
  cv.folds = 5,     #5 fold cross validation
  n.cores = NULL, 
  verbose = FALSE  #don't print progress and performance indicators
  )  
gbm.perf(gbm.fit, plot.it = FALSE,method = "cv")
```

```{r}
hyper_grid <- expand.grid(
  shrinkage = c(.005,0.01, .1),
  interaction.depth = c(1,5,10),
  n.minobsinnode = c(5, 10, 15),   # minimum number of observations in tree terminal nodes
  bag.fraction = c(0.65,0.8,0.9), 
  optimal_trees = 0,               # a place to dump results
  min_RMSE = 0                     # a place to dump results
)


# grid search 
for(i in 1:nrow(hyper_grid)) {
  
  set.seed(66)
  
  gbm.tune <- gbm(
    formula = checklist_pps ~ .,
    distribution = "gaussian",
    data = census_ebird_nor_trn,
    n.trees = 2000,
    interaction.depth = hyper_grid$interaction.depth[i],
    shrinkage = hyper_grid$shrinkage[i],
    n.minobsinnode = hyper_grid$n.minobsinnode[i],
    bag.fraction = hyper_grid$bag.fraction[i],
    train.fraction = .75,
    n.cores = NULL, 
    verbose = FALSE
  )
  
  hyper_grid$optimal_trees[i] <- which.min(gbm.tune$valid.error)
  hyper_grid$min_RMSE[i] <- sqrt(min(gbm.tune$valid.error))
}

hyper_grid %>% 
  dplyr::arrange(min_RMSE) %>%
  head(10)
```

####3.4 Train the final GBM model, identify the most relevant variables
```{r}
set.seed(66)
# train GBM model
gbm.fit.final <- gbm(
  formula = checklist_pps ~ .,
  distribution = "gaussian",
  data = census_ebird_nor_trn,
  n.trees = 1136,
  interaction.depth = 10,
  shrinkage = 0.01,
  n.minobsinnode = 5,
  bag.fraction = .65, 
  train.fraction = .75,
  n.cores = NULL, 
  verbose = FALSE
  )  

par(mar = c(5, 8, 1, 1))
summary(
  gbm.fit.final, 
  cBars = 10,
  method = relative.influence, # also can use permutation.test.gbm
  las = 2
  )

error[2,1] <- sqrt(sum((gbm.fit.final$fit-census_ebird_nor_trn$checklist_pps)^2)/length(census_ebird_nor_trn$checklist_pps))
```

```{r}
fig1=gbm.fit.final %>%
  partial(
    pred.var = "median_ageE", 
    n.trees = gbm.fit.final$n.trees, 
    grid.resolution = 100
    ) %>%
  autoplot(rug = TRUE, train = census_ebird_nor_trn)

fig2=gbm.fit.final %>%
  partial(
    pred.var = "totalSpSeen", 
    n.trees = gbm.fit.final$n.trees, 
    grid.resolution = 100
    ) %>%
  autoplot(rug = TRUE, train = census_ebird_nor_trn)

fig3=gbm.fit.final %>%
  partial(
    pred.var = "black", 
    n.trees = gbm.fit.final$n.trees, 
    grid.resolution = 100
    ) %>%
  autoplot(rug = TRUE, train = census_ebird_nor_trn)

fig4=gbm.fit.final %>%
  partial(
    pred.var = "female", 
    n.trees = gbm.fit.final$n.trees, 
    grid.resolution = 100
    ) %>%
  autoplot(rug = TRUE, train = census_ebird_nor_trn)

fig5=gbm.fit.final %>%
  partial(
    pred.var = "bachelorhigher_p", 
    n.trees = gbm.fit.final$n.trees, 
    grid.resolution = 100
    ) %>%
  autoplot(rug = TRUE, train = census_ebird_nor_trn)

grid.arrange(fig1, fig2,fig3,fig4,fig5, top="Partial dependence plots for the top 5 relevant variables")
```

####3.5 Test the final GBM model on test data
```{r, message=F}
gbm_pred <- predict(gbm.fit.final, n.trees = gbm.fit.final$n.trees, census_ebird_nor_tst)
error[2,2]=RMSE(gbm_pred, census_ebird_nor_tst$checklist_pps)

gbmMod_result <- as.data.frame(cbind(actuals=census_ebird_nor_tst$checklist_pps, predicteds=gbm_pred))

ggplot(gbmMod_result,aes(x=actuals, y=predicteds)) + 
  geom_point(shape=18, color="brown1")+
  geom_smooth(method=lm)+
  labs(title="Relationship between predicted results and actual results",
       x="actual results", y = "predicted results")+
  theme(plot.title = element_text(hjust = 0.5),text = element_text(size=15))

print(error)
```

###Conclusions:
###1. Checklist number per person is a good metric for measuring popularity of bird watching.
###2. Santa Cruz, AZ is a potential place to start the business. People prefer 1-3 hours trips in Spring on Saturday.
###3. Our potential customers are more likely to be older man with higher education level.
